---
title: "Parameters for simulations using `retrocombinator`"
author: "Anindya Sharma"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Running Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Parameters and their defaults

TODO: 
* Import sequences as a starting point instead of a random sequence
* Give a print message that describes the paramaters that have been run in an
  intuitive fashion
* For each object, have a print function that can do this
* By default, print as a message, but give users the option to dump to a log
  file
* Search parameter space and label parameters with what scenarios they are
  likely to lead to


* `SequenceParams` represents what set of retrotransposons the simulation
  starts off with. It starts of multiple identical copies of the same sequence,
  where that sequence is sampled from a uniform A/G/C/T distribution. It is constructed
  from one of the following:
    * `num_seq : numeric` Number of initial retrotransposons **(default =
      100)**
    * `seq_length : numeric` Sequence length (number of nucleotides) of the
      retrotransposons **(default = 5000)**
* `MutationParams` represents how what nucleotide subtstition model will be
  used to modify the sequences during the simulation. It is one of the following
  `character` literals:
    * `"JC69"` Jules and Cantor - 1969
    * `"K80"` Kimura - 1980 **(default)**
    * `"F81"` Felsenstein - 1981
    * `"HKY85"` Hasegawa, Kishino and Yano - 1985
    * `"TN93"` Timura and Nei - 1993
    * `"GTR"`  General Time Reversible Model, Tavaré - 1986
* `FlagParams` represents the parameters used during a simulation in which
  we keep track of a retrotransposon's 'active' status. A retrotransposon that
  is active is capable of transposition or bursting, but potentially loses its
  active status if a point mutation affects a critical part of the sequence.
    * `length_critical_region: numeric`  Sequence length (number of nucleotides)
      of the critical region of a retrotransposon **(default = 10)**
    * `prob_inactivity_when_mutated : numeric` The probability that a point
      mutation to the critical region causes a sequence to become inactive
      **(default = 0.001)**
    * `max_inactive_copies : numeric` What is the largest population size of
      active sequences to keep track of? **(default = 500)** 
* `BurstParams` represents how an active transposon will burst during
  transposition. It comprises of the following:
    * `burst_probability : numeric` The probability that an active
      retrotransposon will increase in copy number during a time jump of one
      timestep **(default = 0.1)**
    * `burst_mean : numeric` The Poisson mean for the distribution that
      specifies how many new sequences an active sequence will create during
      bursting **(default = 1)**
    * `recomb_mean : numeric` The expected number of template switches during 
      recombination between two sequences (chosen from a Poisson distribution
      with this as its mean) **(default = 0.5)**
    * `recomb_similarity : numeric` How similar does an active sequence have to
      be with another sequence for them to be allowed to be recombine during
      transposition? **(default = 0.85)**
* `SimulationParams` represents how long the simulation will run for, and at what
  timescale. It comprises of the following:
    * `num_jumps : numeric` The number of steps (jumps) in our simulation
      **(default = 20)**
    * `timestep : numeric` How much real time does one step (jump) in our
      simulation measure, in millions of years **(default = 1)**
    * `max_active_copies : numeric` What is the largest population size of
      active sequences to keep track of? **(default = 500)**
* `SpeciationParams` represents how we keep track of families during the
  simulation. It comprises of the following:
    * `selection_threshold : numeric` What sequence similarity to the
      original sequence do we wish to maintain? Sequences that diverge beyond
      this similarity level are dropped over the course of simulation
      **(default = 0.5)**
    * `family_similarity : numeric` What sequence similarity do two sequences
      have to be to each other for them to be considered to be of the same
      family? **(default = 0.7)**
    * `family_coherence : numeric` What proportion of the overall sequence
      similarity matrix of a family needs to score be above `family_similarity`
      before we decide the family has split into further sub-families?
      **(default = 0.5)**
* `OutputParams` represents how and where the output of the simulation will
  be saved. It comprises of the following:
    * `file_out : character` Where should the simulation be saved? **(default =
      'file.out')**
    * `num_out_seqs : numeric` How many times during the event of the simulation
      should we output the raw sequences (as nucleotide strings) themselves?
      **(default = 2)**
    * `num_out_init : numeric` How many times during the event of the simulation
      should we output the distance of sequences to the initial sequence?
      **(default = 10)**
    * `num_out_pair : numeric` How many times during the event of the simulation
      should we output pairwise distances between all pairs of sequences?
      **(default = 10)**
    * `num_out_fam : numeric` How many times during the event of the simulation
      should we output the family tags of each of the sequences?
      **(default = 10)**
* `SeedParams` represents how to select the seed for randomisation for the
  simulation. It comprises of the following:
    * `to_randomise : logical` Should this simulation be run with a random seed
      to begin with? (The seed is based on system time) **(default = false)**
    * `to_seed : logical` Should this simulation be run with a specified seed to
      begin with? (This overwrites `to_randomise`) **(default = true)**
    * `seed : numeric` If `to_seed` is `TRUE`, what should the initial seed for
      the random number generator be? **(default = 0)**

# How to use

To run a simulation, just call the main function `simulate()` and it will run with the default
parameters. 

```{r, eval = FALSE}
simulate()
```

To overwrite any parameters, first create objects to represent the parameters
you wish to overwrite. Say for example you want to change the recombination
mean, and the flagging parameters, run the following code. Whenever a parameter is
not specified explicitly, the default will be used.

```{r, eval = FALSE}
burst_params <- BurstParams(recomb_mean = 1.0)
flag_params <- FlagParameters(length_critical_region = 0,
                              prob_inactivity_when_mutated = 0)
simulate(BurstParams = burst_params, FlagParams = flag_params)
```
