---
title: "Running Simulations for Evolution of Transposons"
author: "Anindya Sharma"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `retrocombinator` package provides a method for simulating the molecular
evolutionary process of retrotransposon recombination over large scales of time.

To use it, load the library first.
```{r}
library(retrocombinator)
```

Simulations can be done in two ways - with and without flags for inactivity.
The functions can be called in the following way:

```{r}
# Calling a simulation without flags
simulate_without_flags_random(num_seq = 5, seq_length = 20,
                       point_mutation_model = "K80",
                       num_jumps = 20, timestep = 2,
                       burst_probability = 0.1, burst_mean = 1,
                       recomb_mean = 1,
                       file_out = "without_flags_vignette.out",
                       num_out_init = 10, num_out_tags = 10,
                       num_out_seqs = 10, num_out_pair = 10,
                       to_seed = TRUE, seed = 0)
```

TODO: Add a not about how `num_out_pair` needs to be a multiple of
`num_out_init` for helpful plotting
This data can then be analysed in the following way:

```{r}
# Import the data file
data_without_flags <- input_data("without_flags_vignette.out")
```

This data object can be inspected to find the tags for each sequence, their distance to the initial sequence, and the pairwise distances between sequences.

```{r}
head(data_without_flags$init)
head(data_without_flags$pair)
```

Similarly, we can run a simulation with flags.

```{r}
# Calling a simulation without flags
simulate_with_flags_random(num_seq = 10, 
                           seq_length = 200,
                    point_mutation_model = "K80",
                    num_sensitive_posns = 10, 
                    inactive_probability = 0.1, 
                    num_jumps = 50, 
                    timestep = 2,
                    burst_probability = 0.1, 
                    burst_mean = 1,
                    recomb_mean = 1,
                    selection_threshold = 60, 
                    fam_proportion = 0.7, 
                    fam_percentage = 70,
                    file_out = "with_flags_vignette.out",
                    num_out_init = 10, 
                    num_out_tags = 10,
                    num_out_seqs = 10, 
                    num_out_pair = 10,
                    to_seed = TRUE, 
                    seed = 0)
```

Again the results of the simulation can be analysed using the `input_data` function.

```{r}
data_with_flags <- input_data("with_flags_vignette.out")
head(data_with_flags$init)
head(data_with_flags$pair)
```

Visualise the results
```{r}
plot_initial_distances(data_with_flags)
plot_family_distances(data_with_flags)
```
