---
title: "Running simulations using `retrocombinator`"
author: "Anindya Sharma"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Running Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Running Simulations

The `retrocombinator` package provides a method for simulating the molecular
evolutionary process of retrotransposon recombination over large scales of time.

To use it, load the library first.
```{r}
library(retrocombinator)
```

To run a simulation, call the `simulateEvolution()` function. This runs a
simulation with default parameters, and returns a path to a simulation output
file.

```{r}
simulateEvolution()
```

To overwrite any parameters, first create objects to represent the parameters
you wish to overwrite, such as in the example given below. Refer to the section
below for what the various parameters are.  Whenever a parameter is not
specified explicitly, the default will be used.

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---
burstParams <- BurstParams(recombMean = 1.0)
flagParams <- FlagParams(lengthCriticalRegion = 5,
                         probInactiveWhenMutated = 0.1)
simulateEvolution(burstParams = burstParams, flagParams = flagParams)
# ----------------------------------------
```

### Using custom sequences

By default, `simulateEvolution()` runs a simulation on randomly generated
sequences. To use a custom sequence, specify it in `SequenceParams` (more
details in the section below, that describes all possible parameters to the
simulation).

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---

# Obtain your sequence in your favourite way
library(Biostrings)
fastaInput <- readDNAStringSet('path/to/your/FASTA/file')
yourSequence <- toString(fastaInput$yourSequence)

# Override the sequence parameters and pass it to the simulation
sequenceParams <- SequenceParams(initialSequence = yourSequence)
simulateEvolution(sequenceParams = sequenceParams)
# ----------------------------------------
```
  
## Simulation parameters and their defaults

* `SequenceParams` represents what set of retrotransposons the simulation starts
  off with. It is constructed by either specifying an explicit initial
  sequence, or a sequence length (in which case the sequence is created randomly
  by sampling from a uniform A/G/C/T distribution.) Besides that, the number of
  initial copies of the initial retrotransposon needs to be specified (the
  simulation creates multiple copies of the initial sequence to begin with):
    * `sequenceLength : numeric` Sequence length (number of nucleotides) of the
      initial retrotransposon (constructed randomly); ignored if `initialSequence` 
      is not `NULL` 
      **(default = 5000)**
    * `initialSequence : character` The initial retrotransposon to consider; can
      be `NULL` if `sequenceLength` is specified instead (**default = `NULL`**)
    * `numInitialCopies : numeric` Number of copies of initial retrotransposon **(default = 20)**
* `MutationParams` represents how what nucleotide subtstition model will be
  used to modify the sequences during the simulation. It is constructed from one
  of the following `character` literals to the argument `model`:
    * `"JC69"` Jules and Cantor - 1969
    * `"K80"` Kimura - 1980 **(default)**
    * `"F81"` Felsenstein - 1981
    * `"HKY85"` Hasegawa, Kishino and Yano - 1985
    * `"TN93"` Timura and Nei - 1993
    * `"GTR"`  General Time Reversible Model, Tavaré - 1986
* `FlagParams` represents the parameters used during a simulation in which
  we keep track of a retrotransposon's *active* status. A retrotransposon that
  is active is capable of transposition or bursting, but potentially loses its
  active status if a point mutation affects a critical part of the sequence.
    * `lengthCriticalRegion: numeric`  Sequence length (number of nucleotides)
      of the critical region of a retrotransposon **(default = 10)**
    * `probInactiveWhenMutated : numeric` The probability that a point
      mutation to the critical region causes a sequence to become inactive
      **(default = 0.01)**
    * `maxInactiveCopies : numeric` The largest population size of
      inactive sequences to keep track of (if this is exceeded, sequences are
      randomly discarded to simulate death) **(default = 500)** 
* `SimulationParams` represents how long the simulation will run for, and at what
  timescale. It comprises of the following:
    * `numSteps : numeric` The number of steps in our simulation - the
      simulation starts at a step count of 0, and runs until the step count is `numSteps`
      **(default = 20)**
    * `timestep : numeric` How much real time does one step in our
      simulation measure, in millions of years **(default = 1)**
    * `maxActiveCopies : numeric` The largest population size of
      active sequences to keep track of (if this is exceeded, sequences are
      randomly discarded to simulate death) **(default = 500)** 
* `BurstParams` represents how an active retrotransposon will burst during
  transposition. It comprises of the following:
    * `burstProbability : numeric` The probability that an active
      retrotransposon will increase in copy number during a time jump of one
      timestep **(default = 0.1)**
    * `burstMean : numeric` The Poisson mean for the distribution that
      specifies how many new sequences an active sequence will create during
      bursting **(default = 1)**
    * `recombMean : numeric` The expected number of template switches during 
      recombination between two sequences (chosen from a Poisson distribution
      with this as its mean) **(default = 1.5)**
    * `recombSimilarity : numeric` How similar does an active sequence have to
      be with another sequence for them to be allowed to be recombine during
      transposition? **(default = 0.85)**
* `SpeciationParams` represents how we keep track of species during the
  simulation. It comprises of the following:
    * `selectionThreshold : numeric` What sequence similarity to the
      original sequence do we wish to maintain? Sequences that diverge beyond
      this similarity level are dropped over the course of simulation
      **(default = 0.5)**
    * `speciesSimilarity : numeric` What sequence similarity do two sequences
      have to be to each other for them to be considered to be of the same
      species? **(default = 0.9)**
    * `speciesCoherence : numeric` What proportion of the overall sequence
      similarity matrix of a species needs to score be above `speciesSimilarity`
      before we decide the species has split into two species?
      **(default = 0.8)**
* `OutputParams` represents how and where the output of the simulation will
  be saved. It comprises of the following:
    * `outputFilename : character` Where should the simulation be saved? **(default =
      'simulationOutput.out')**
    * `outputNumRawSequence : numeric` How many times during the event of the simulation
      should we output the raw sequences (as nucleotide strings) themselves?
      **(default = 2)**
    * `outputNumInitialDistance : numeric` How many times during the event of the simulation
      should we output the distance of sequences to the initial sequence?
      **(default = 10)**
    * `outputNumPairwiseDistance : numeric` How many times during the event of the simulation
      should we output pairwise distances between all pairs of sequences?
      **(default = 10)**
    * `outputNumSpeciesTags : numeric` How many times during the event of the simulation
      should we output the species tags of each of the sequences?
      **(default = 10)**
* `SeedParams` represents how to select the seed for randomisation for the
  simulation. It comprises of the following:
    * `toSeed : logical` Should this simulation be run with a specified seed to
      begin with? If `FALSE`, a seed based on system time is used **(default =
      true)**
    * `seedForRandom : numeric` If `toSeed` is `TRUE`, what should the initial seed for
      the random number generator be? **(default = 0)**

## Output

The simulation creates a custom output file and returns a path to it (specified
by `outputFilename` in `OutputParams`, which by default is `'simulationOutput.out'`). 
This file can be parsed by `parseSimulationOutput()`.

```{r}
data <- parseSimulationOutput('simulationOutput.out')

```

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---

# Alternatively, using the pipe operator
library(magrittr) # For %>%
data <- simulateEvolution() %>%
  parseSimulationOutput()

# ----------------------------------------
```

This contains a list of two dataframes, that keep track of retrotransposon
sequences, their similarity to the initial sequence, their pairwise similarity,
and their lineage, across the timesteps of the evolution. 

For pairwise similarity, pairs are not duplicated, and only distinct pairs are
given. That is, for any timestep and sequences with `sequenceId` $a$ and $b$
present at that time, a row is present in the `pairwise` dataframe if and only
if $a < b$ .

```{r}
print(colnames(data$sequences))

# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# sequenceId - the unique ID of the sequence (to track it over time); initial
#              sequence has sequenceId 0 to (numInitialCopies-1)
# speciesId - the unique ID of a species; initial species has speciesId 0
# previousSpeciesId - the unique ID of the species this species emerged from; 
#                     (-1 if nothing)
# parentMain - the unique ID of the sequence this burst from; 
#              (-1 if nothing)
# parentOther - the unique ID of the sequence its parent recombined with; 
#               (-1 if nothing)
# distanceToInitial - the distance to the initial sequence
# rawSequence - the raw sequence, as a character
# isActive - whether or not the sequence is capable of transposition

print(colnames(data$pairwise))

# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# sequenceId1 - an ID of a sequence present at the time
# sequenceId2 - an ID of a different sequence present at the time; not all pairs
#               are given - that is, for sequences a and b, either (a, b) or (b, a) 
#               is present as a row but not both
# distancePairwise - the distance between the two sequences

# TODO: colnames(data$params)
```

## Analysis

Summary statistics can be obtained from the data as follows.

```{r}
# TODO: summariseEvolution(data)
```

The data can be visualised using the `plotEvolution` function
```{r}
plotEvolution(data, "initial")   # Distance to initial sequence
plotEvolution(data, "pairwise")  # Pairwise distances between sequences
# TODO: plotEvolution(data, "phylogeny") # Phylogeny of the sequences
```
