---
title: "Running simulations using `retrocombinator`"
author: "Anindya Sharma"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Running Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Running Simulations

The `retrocombinator` package provides a method for simulating the molecular
evolutionary process of retrotransposon recombination over large scales of time.

To use it, load the library first.
```{r}
library(retrocombinator)
```

To run a simulation, call the `simulateEvolution()` function. This runs a
simulation with default parameters, and returns a path to a simulation output
file.

```{r}
simulateEvolution()
```

To overwrite any parameters, first create objects to represent the parameters
you wish to overwrite, such as in the example given below. Refer to the section
below for what the various parameters are.  Whenever a parameter is not
specified explicitly, the default will be used.

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---
recombParams <- RecombParams(recombMean = 1.0)
activityParams <- ActivityParams(lengthCriticalRegion = 20,
                                 probInactiveWhenMutated = 0.1)
simulateEvolution(recombParams = recombParams, activityParams = activityParams)
# ----------------------------------------
```

### Using custom sequences

By default, `simulateEvolution()` runs a simulation on randomly generated
sequences. To use a custom sequence, specify it in `SequenceParams` (more
details in the section below, that describes all possible parameters to the
simulation).

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---

# Obtain your sequence in your favourite way
library(Biostrings)
fastaInput <- readDNAStringSet('path/to/your/FASTA/file')
yourSequence <- toString(fastaInput$yourSequence)

# Override the sequence parameters and pass it to the simulation
sequenceParams <- SequenceParams(initialSequence = yourSequence)
simulateEvolution(sequenceParams = sequenceParams)
# ----------------------------------------
```
  
## Simulation parameters and their defaults

* `SequenceParams` represents what set of retrotransposons the simulation starts
  off with. It is constructed by either specifying an explicit initial
  sequence, or a sequence length (in which case the sequence is created randomly
  by sampling from a uniform A/G/C/T distribution.) Besides that, the number of
  initial copies of the initial retrotransposon needs to be specified (the
  simulation creates multiple copies of the initial sequence to begin with):
    * `sequenceLength : numeric` Sequence length (number of nucleotides) of the
      initial retrotransposon (constructed randomly); ignored if `initialSequence` 
      is not `NULL` 
      **(default = 5000)**
    * `initialSequence : character` The initial retrotransposon to consider; can
      be `NULL` if `sequenceLength` is specified instead (**default = `NULL`**)
    * `numInitialCopies : numeric` Number of copies of initial retrotransposon **(default = 20)**
* `ActivityParams` represents the parameters used during a simulation in which
  we keep track of a retrotransposon's *active* status. A retrotransposon that
  is active is capable of transposition or bursting, but potentially loses its
  active status if a point mutation affects a critical part of the sequence.
    * `lengthCriticalRegion: numeric`  Sequence length (number of nucleotides)
      of the critical region of a retrotransposon **(default = 10)**
    * `probInactiveWhenMutated : numeric` The probability that a singe point
      mutation to the critical region causes a sequence to become inactive
      **(default = 0.01)**
* `MutationParams` represents how what nucleotide substitution model will be
  used to modify the sequences during the simulation. It is constructed from one
  of the following `character` literals to the argument `model`, and the
  appropriate optional arguments (all of type `numeric`) to set the parameters
  within the model.
    * `"GTR"`  General Time Reversible Model, Tavaré - 1986 **(this model is not
      fully supported)**
      * `pi_T`, `pi_C`, `pi_A`, `pi_G` - proportion of the sequence that
        comprises of the nucleotide T, C, A or G (respectively )after the system
        has reached equilibrium
      * `T2C`, `T2A`, `T2G`, `C2A`, `C2G`, `A2G` - X2Y represents the
        corresponding entry in the transition matrix, which is the rate at which
        nucleotide X mutates into nucleotide Y
      * `scale` - the scalar by which the default transition matrix is to be
        scaled by in order to get the number of expected mutations per
        millions of years
    * `"TN93"` Timura and Nei - 1993
      * `pi_T`, `pi_C`, `pi_A`, `pi_G` - proportion of the sequence that
        comprises of the nucleotide T, C, A or G (respectively )after the system
        has reached equilibrium
      * `k1` - rate of transition from T to C (or vice-versa) when the
        transversion rate is 1
      * `k2` - rate of transition from A to G (or vice-versa) when the
        transversion rate is 1
      * `scale` - the scalar by which the default transition matrix is to be
        scaled by in order to get the number of expected mutations per
        millions of years
    * `"HKY85"` Hasegawa, Kishino and Yano - 1985
      * `pi_T`, `pi_C`, `pi_A`, `pi_G` - proportion of the sequence that
        comprises of the nucleotide T, C, A or G (respectively )after the system
        has reached equilibrium
      * `k` - rate of any transition (i.e. either a swap between T and C or a
        swap between A and G) when the transversion rate is 1
      * `scale` - the scalar by which the default transition matrix is to be
        scaled by in order to get the number of expected mutations per
        millions of years
    * `"F81"` Felsenstein - 1981
      * `pi_T`, `pi_C`, `pi_A`, `pi_G` - proportion of the sequence that
        comprises of the nucleotide T, C, A or G (respectively )after the system
        has reached equilibrium
      * `scale` - the scalar by which the default transition matrix is to be
        scaled by in order to get the number of expected mutations per
        millions of years
    * `"K80"` Kimura - 1980 **(default)**
      * `k` - rate of any transition (i.e. either a swap between T and C or a
        swap between A and G) when the transversion rate is 1
      * `scale` - the scalar by which the default transition matrix is to be
        scaled by in order to get the number of expected mutations per
        millions of years
    * `"JC69"` Jules and Cantor - 1969
        * `scale` - the scalar by which the default transition matrix is to be
          scaled by in order to get the number of expected mutations per
          millions of years
* `BurstParams` represents how an active retrotransposon will burst during
  transposition. It comprises of the following:
    * `burstProbability : numeric` The probability that an active
      retrotransposon will increase in copy number during a time jump of one
      timestep **(default = 0.1)**
    * `burstMean : numeric` The Poisson mean for the distribution that
      specifies how many new sequences an active sequence will create during
      bursting **(default = 1)**
    * `maxTotalCopies : numeric` The largest population size of sequences to
      keep track of (if this is exceeded, sequences are randomly discarded to
      simulate death) **(default = 50)** 
* `RecombParams` represents how two transposons recombine during transposition
    * `recombMean : numeric` The expected number of template switches during 
      recombination between two sequences (chosen from a Poisson distribution
      with this as its mean) **(default = 1.5)**
    * `recombSimilarity : numeric` How similar does an active sequence have to
      be with another sequence for them to be allowed to be recombine during
      transposition? **(default = 0.80)**
* `SelectionParams` represents what sequences we are interested in storing during
  the simulation. It comprises of the following:
    * `selectionThreshold : numeric` What sequence similarity to the
      original sequence do we wish to maintain? Sequences that diverge beyond
      this similarity level are dropped over the course of simulation
      **(default = 0.3)**
* `FamilyParams` represents how we keep track of families during the
  simulation. It comprises of the following:
    * `familyCoherence : numeric` What sequence similarity do two sequences
      have to be to each other for them to be considered to be of the same
      family? **(default = 0.70)**
    * `maxFamilyRepresentatives : numeric` How many family representatives to
      keep track of during the simulation?  **(default = 20)**
* `SimulationParams` represents how long the simulation will run for, and at what
  timescale. It comprises of the following:
    * `numSteps : numeric` The number of steps in our simulation - the
      simulation starts at a step count of 0, and runs until the step count is `numSteps`
      **(default = 20)**
    * `timePerStep : numeric` How much real time does one step in our
      simulation measure, in millions of years **(default = 1)**
* `OutputParams` represents how and where the output of the simulation will
  be saved. It comprises of the following:
    * `outputFilename : character` Where should the simulation be saved? **(default =
      'simulationOutput.out')**
    * `outputNumInitialDistance : numeric` How many times during the simulation
      should we output the distance of sequences to the initial sequence?
      **(default = 10)**
    * `outputNumPairwiseDistance : numeric` How many times during the simulation
      should we output pairwise distances between all pairs of sequences?
      **(default = 10)**
    * `outputNumFamilyLabels : numeric` How many times during the simulation
      should we output the family numbers of each sequence?
      **(default = 10)**
    * `outputNumFamilyMatrix : numeric` How many times during the simulation
      should we output the pairwise distances between family representatives?
      **(default = 10)**
    * `outputMinSimilarity : numeric` What is the minimum similarity
      between two sequences we should report on? **(default = 0.5)**
* `SeedParams` represents how to select the seed for randomisation for the
  simulation. It comprises of the following:
    * `toSeed : logical` Should this simulation be run with a specified seed to
      begin with? If `FALSE`, a seed based on system time is used **(default =
      `TRUE`)**
    * `seedForRNG : numeric` If `toSeed` is `TRUE`, what should the initial seed for
      the random number generator be? **(default = 0)**

## Output

The simulation creates a custom output file and returns a path to it (specified
by `outputFilename` in `OutputParams`, which by default is `'simulationOutput.out'`). 
This file can be parsed by `parseSimulationOutput()`.

```{r}
data <- parseSimulationOutput('simulationOutput.out')

```

```{r, eval = FALSE}
# --- NOT RUN (this is an alternative) ---

# Alternatively, using the pipe operator
library(magrittr) # For %>%
data <- simulateEvolution() %>%
  parseSimulationOutput()

# ----------------------------------------
```

This contains

* a list of the parameters used to run the simulation
* a data frame of the sequences over time and their similarity to the initial
  sequence
* a data frame of the pairwise similarity of sequences over time
* a data frame of the family representatives for the retrotransposons over time
* a data frame of the pairwise similarity of the family representatives for the
  retrotransposons over time

For pairwise similarity, pairs are not duplicated, and only distinct pairs are
given. That is, for any timestep and sequences with `sequenceId` $a$ and $b$
present at that time, a row is present in the `pairwise` dataframe if and only
if $a < b$ .

```{r}
print(names(data$params))
# Prints out all the parameters the simulation was run with

print(colnames(data$sequences))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# sequenceId - the unique ID of the sequence (to track it over time); initial
#              sequence has sequenceId 0 to (numInitialCopies-1)
# parentMain - the unique ID of the sequence this burst from; 
#              (-1 if nothing)
# parentOther - the unique ID of the sequence its parent recombined with; 
#               (-1 if nothing)
# distanceToInitial - the distance to the initial sequence
# isActive - whether or not the sequence is capable of transposition

print(colnames(data$pairwise))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# sequenceId1 - an ID of a sequence present at the time
# sequenceId2 - an ID of a different sequence present at the time; not all pairs
#               are given - that is, for sequences a and b, either (a, b) or (b, a) 
#               is present as a row but not both
# distancePairwise - the distance between the two sequences

print(colnames(data$familyRepresentatives))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# familyId - the unique ID of the family representative (to track it over time)
# creationTime - time this family representative was created (in millions of years)
# sequenceId - a sequence belonging to that family (sequences belonging to a
#              family are listed as separate rows, and all sequences belonging
#              to that family are listed)

print(colnames(data$familyPairwise))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# familyId1 - an ID of a family representative present at the time
# familyId2 - an ID of a different family representative present at the time;
#             not all pairs are given - that is, for sequences a and b, either (a, b) or (b,
#             a) is present as a row but not both
# distancePairwise - the distance between the two family representatives
```

## Visualising the results

The results of the simulation can be visualised using the `plotEvolution`
function.

```{r}
plotEvolution(data, "initial")  # Similarity to initial sequence
plotEvolution(data, "pairwise") # Pairwise similarities between sequences
plotEvolution(data, "families") # Families of sequences
```

## Summarising the results

The results of the simulation can also be visualised using the
`summariseEvolution` function, which can in turn be visualised using the
`plotSummary` function.

```{r}
summary <- summariseEvolution(data, 'initial')

print(colnames(summary))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# numSequences - how many sequences are present at that time?
# divergenceMin - the minimum similarity to the initial sequence
# divergenceQ25 - the first quartile similarity to the initial sequence
# divergenceMedian - the median similarity to the initial sequence
# divergenceQ75 - the third quartile similarity to the initial sequence
# divergenceMax - the maximum quartile similarity to the initial sequence
# divergenceMean - the mean similarity to the initial sequence
```

```{r}
summary <- summariseEvolution(data, 'pairwise')

print(colnames(summary))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# pairwiseMin - the minimum pairwise similarity between sequences
# pairwiseQ25 - the first quartile pairwise similarity between sequences
# pairwiseMedian - the median pairwise similarity between sequences
# pairwiseQ75 - the third quartile pairwise similarity between sequences
# pairwiseMax - the maximum pairwise similarity between sequences
# pairwiseMean - the mean similarity to the initial sequence
```

```{r}
summary <- summariseEvolution(data, 'families')

print(colnames(summary))
# step - timestep in the simulation
# realTime - time since the start of the simulation (in millions of years)
# numSequences - how many sequences are present at that time?
# numFamilies - how many families are present at that time?
# familiesMin - the minimum family size
# familiesQ25 - the first quartile family size
# familiesMedian - the median family size
# familiesQ75 - the third quartile family size
# familiesMax - the maximum quartile family size
# familiesMean - the mean family size
```

```{r, eval = FALSE}
summary <- summariseEvolution(data, 'history')

# TODO: This section
summary$timeOfDeath
# If NA, the sequences did not die.
# Else, the realTime (in millions of years) at when the population became 0
summary$timeBetweenFamilies
# The min, Q25, median, Q75 and max of the time between new families emerging
# (in millions of years)
summary$timeBetweenResurrection
```

## Example with all parameters

```{r}
sequenceParams <- SequenceParams(initialSequence = "TCAGTCAGTCAGTCAGTGTG",
                                 numInitialCopies = 10)
activityParams <- ActivityParams(lengthCriticalRegion = 2,
                                 probInactiveWhenMutated = 0.1)
mutationParams <- MutationParams(model = "F81")
burstParams <- BurstParams(burstProbability = 0.2,
                           burstMean = 2,
                           maxTotalCopies = 40)
recombParams <- RecombParams(recombMean = 2.0,
                             recombSimilarity = 0.85)
selectionParams <- SelectionParams(selectionThreshold = 0.25)
familyParams <- FamilyParams(familyCoherence = 0.65,
                             maxFamilyRepresentatives = 15)
simulationParams <- SimulationParams(numSteps = 35,
                                     timePerStep = 1.5)
outputParams <- OutputParams(outputFilename = 'simulationExampleParameters.out',
                             outputNumInitialDistance = 5,
                             outputNumPairwiseDistance = 5,
                             outputNumFamilyLabels = 5,
                             outputNumFamilyMatrix = 5,
                             outputMinSimilarity = 0.1)
seedParams <- SeedParams(toSeed = TRUE, seedForRNG = 1)
simulateEvolution(sequenceParams = sequenceParams,
                  activityParams = activityParams,
                  mutationParams = mutationParams,
                  burstParams = burstParams,
                  recombParams = recombParams,
                  selectionParams = selectionParams,
                  familyParams = familyParams,
                  simulationParams = simulationParams,
                  outputParams = outputParams)
```

